---
title: "Inner West LGA Sales and Rent Report"
date: "`r Sys.Date()`"
author: "Cooper Gee"
format: 
  pdf:
    documentclass: scrreprt
    pdf-engine: pdflatex
    echo: false
table-of-contents: true
number-sections: true
appendix-style: default
bibliography: bibliography.bib
csl: ieee.csl
---

```{r setup}
#| output: false

library(tidyverse)
library(readxl)
library(ggpubr)

load("data/rdata/sales_df.RData")
load("data/rdata/rents_df.RData")
load("data/rdata/rba_cash_rate.RData")


innerwest_colour = "#32a852"
greatersyd_colour = "#e82b09"
wholensw_colour = "#099ee8"

inner_west_sales <- sales_df |> dplyr::filter(lga == "Inner West" &
                                              greater_sydney == "Total" &
                                              gmr == "Total" &
                                              rings == "Total")
inner_west_rents <- rents_df |> dplyr::filter(lga == "Inner West" &
                                              greater_sydney == "Total"&
                                              gmr == "Total"&
                                              rings == "Total")

nsw_sales <- sales_df |> dplyr::filter(lga == "Total",
                                       greater_sydney == "Total",
                                       gmr == "Total",
                                       rings == "Total")
nsw_rents <- rents_df |> dplyr::filter(lga == "Total",
                                       greater_sydney == "Total",
                                       gmr == "Total",
                                       rings == "Total")

sydney_sales <- sales_df |> dplyr::filter(lga == "Total",
                                       greater_sydney == "Greater Sydney",
                                       gmr == "Total",
                                       rings == "Total")
sydney_rents <- rents_df |> dplyr::filter(lga == "Total",
                                       greater_sydney == "Greater Sydney",
                                       gmr == "Total",
                                       rings == "Total")


```

## Introduction

The property market in Australia is a topic of much debate recently. There has been commentary regarding the cost of rent and the going price of a home. What this report sets out to examine is how the rental and property market in the Inner West local government area (LGA) of Sydney, New South Wales has changed over the period of January 2018 to June 2024. In doing so, this report has also provided insights into the broader property and rental market of Greater Sydney and New South Wales.

This time period was chosen for a few reasons. The COVID-19 pandemic occurred in early 2020 to approximately late 2021. In that period, interest rates hit historic lows. In the period after, interest rates rose to combat rising inflation, a phenomenon not local to Australia, but worldwide. With that, recent commentary as of October 2024 has surrounded the rising cost of living in Australia, the apparent lack of affordability of housing, and the impact of property investment in the market.

By using the Inner West LGA as a choice case study of metropolitan Sydney, this report hopes to provide some insights regarding how sales and rents have changed over the prescribed 6 year period, and attribute some reasons for why these changes have occurred.

### The Data

The data used for this report was sourced from the Department of Communities and Justice, a New South Wales State government department @dcj-rents-and-sales. They provide a quarterly report on rents and sales in NSW, with a sheet labelled by post code, and a sheet labelled by local government area. The data used in this report consists of the quarterly report starting from March 2018 to March 2024 (for sales) and June 2024 (for rents).

Reserve Bank of Australia (RBA) cash rate data was sourced from the RBA website @rba-cash-rate. Total average weekly wage in NSW was sourced from the Australian Bureau of Statistics (ABS) @abs-wages.

### The Inner West Local Government Area

The Inner West LGA covers an area of 35 square kilometres in the city of Sydney @council-profile, with an estimated population in 2021 of 182,818 people. The estimated median age of the LGA is 38 years old @census-inner-west. An estimated 48.2% of people aged 15 and over had completed a Bachelor Degree or higher, compared to the national 26.3%. In 2021, there were an estimated 85,118 private dwellings in the LGA, with 74,091 of those dwellings occupied and 8,152 unoccupied.

### Data Loading and Cleaning

The rent and sales data utilised in this report was sourced from the quarterly reports published by the NSW Department of Communities and Justice. The report is provided for rents/sales for each LGA and each postcode.

## Rent in the Inner West LGA

```{r}
#| label: fig-iw-rent
#| fig-cap: "Median weekly rent in the Inner West LGA from January 2018 to July 2024."

set_breaks = scales::breaks_width("6 months")
scale = set_breaks(as.Date(c("2017-04-01","2024-07-01"))) - days(1)

fig_total_rent <- inner_west_rents |> 
  dplyr::filter(dwelling_type == "Total"& no_of_bedrooms == "Total") |>
  ggplot() +
  aes(x = date, y = median_rent_price) +
  geom_point(colour = innerwest_colour) +
  geom_line(colour = innerwest_colour,linetype = "dashed") +
  scale_x_date(breaks = scale,date_labels = "%b %Y") +
  scale_y_continuous(labels=scales::dollar_format()) +
  geom_vline(xintercept=as.Date("2020-03-15"),linetype="dotted") +
  geom_vline(xintercept=as.Date("2021-10-11"),linetype="dotted") +
  annotate("text",label="Pandemic\nPeriod",x=as.Date("2021-01-01"),y=650) +
  theme(axis.text.x=element_text(angle = 45, vjust=1,hjust=1),plot.title = element_text(hjust = 0.5)) +
  labs(x = element_blank(),y = "Median Rent",title = "Median Rent of the Inner West LGA")
fig_total_rent
```

Median rent in the Inner West LGA was trending downwards from January 2018 to January 2020, as shown in @fig-iw-rent. In the quarter ending March 2020 (the first quarter ending within the pandemic period), median rents were at \$555 per week. In the quarter ending June 2021, the lowest median weekly rent in the period is recorded at \$490 per week. This quarter marks a steady trend upwards, barring the March and June quarters of 2023 where the median weekly rent held steady, with the quarter ending June 2024 recording a median weekly rent of \$725. Between that minima recorded in June 2021 and the latest reported quarter in June 2024, median rents have rose by \$235.

### Median Rent by Number of Rooms and Dwelling Type

When considering the number of bedrooms a dwelling has, those with more bedrooms held a consistently higher median rent than those with less bedrooms. When factoring by dwelling type, the median rent for units has been on a relatively steady upward trend post pandemic. The median rents for houses and townhouses follow a similar trend.

```{r}

df <- inner_west_rents |> 
  dplyr::filter(dwelling_type != "Total" & no_of_bedrooms == "Total")
df$dwelling_type <- factor(df$dwelling_type,levels = c("House","Townhouse","Flat/Unit","Other"))

fig_rent_type <- df |>
  ggplot() +
  aes(x = date, y = median_rent_price, colour = dwelling_type) +
  geom_line() +
  scale_x_date(breaks = scale,date_labels = "%b %Y") +
  scale_y_continuous(labels=scales::dollar_format()) +
  geom_vline(xintercept=as.Date("2020-03-15"),linetype="dotted") +
  geom_vline(xintercept=as.Date("2021-10-11"),linetype="dotted") +
  annotate("text",label="Pandemic\nPeriod",x=as.Date("2021-01-01"),y=850) +
  theme(axis.text.x=element_text(angle = 45, vjust=1,hjust=1),
        plot.title = element_text(hjust = 0.5),
        legend.position = "bottom",
        legend.key = element_rect(fill = "white")) +
  labs(x = element_blank(),y = "Median Rent",
       title = "Median Rent of the Inner West LGA\n(Factored by Dwelling Type)",
       colour=element_blank())

fig_rent_type
```

## Rental Bonds in the Inner West

```{r}
#| label: fig-iw-total-bonds
#| fig-cap: "Total number of rental bonds in the Inner West over the reporting period."
fig_iw_bonds_new = inner_west_rents |>
  dplyr::filter(dwelling_type == "Total" & no_of_bedrooms == "Total") |>
  ggplot() +
  aes(x = date, y = no_of_bonds_total) +
  scale_x_date(breaks = scale,date_labels = "%b %Y") +
  scale_y_continuous(labels=scales::comma) +
  geom_vline(xintercept=as.Date("2020-03-15"),linetype="dotted") +
  geom_vline(xintercept=as.Date("2021-10-11"),linetype="dotted") +
  annotate("text",label="Pandemic\nPeriod",x=as.Date("2021-01-01"),y=33000) +
  theme(axis.text.x=element_text(angle = 45, vjust=1,hjust=1),
    plot.title = element_text(hjust = 0.5)) +
  labs(x = element_blank(), y = element_blank(),
       title = "Total Number of Rental Bonds in the Inner West") +
  geom_point(colour = innerwest_colour,shape = "circle") +
  geom_line(linetype = "dashed",colour = innerwest_colour)

fig_iw_bonds_new
```

Looking at @fig-iw-total-bonds, there is a noticeable downturn in total rental bonds in the Inner West from the quarter ending March 2023 onwards. The number of bonds essentially represents the number of active rental properties in the reporting period.

From March 2023, there are three quarterly dips in total bonds, followed by a single hike followed again by another dip. That dip reported in the quarter ending June 2024 is the single largest quarterly loss of bonds post-pandemic, with a loss of 481 bonds. The downward trend realised between March 2023 and June 2024 is noteworthy in that no where else in the reporting period are two consecutive drops in rental stock reported, including the pandemic. Overall between the quarter ending March 2023 and quarter ending June 2024, 863 bonds were wiped from the Inner West rental market.

Consider the reasons why total bonds would shrink. Properties may have been taken off the rental market altogether (sold to owner-occupiers, renovations, sold to developers), or properties may not have been able to find renters.

```{r}
fig_comp <- rents_df |>
  dplyr::filter(
      lga == "Total" & greater_sydney == "Greater Sydney" &
      gmr == "Total" & rings == "Total" & 
      dwelling_type == "Total" & no_of_bedrooms == "Total"
  ) |> 
  ggplot() +
  aes(x = date,y=no_of_bonds_total) +
  geom_point(colour = greatersyd_colour) +
  geom_line(linetype = "dashed",colour = greatersyd_colour) +
  scale_x_date(breaks = scale ,date_labels = "%b %Y") +
  scale_y_continuous(labels=scales::comma) +
  geom_vline(xintercept=as.Date("2020-03-15"),linetype="dotted") +
  geom_vline(xintercept=as.Date("2021-10-11"),linetype="dotted") +
  annotate("text",label="Pandemic\nPeriod",x=as.Date("2021-01-01"),y=610000) +
  scale_color_manual(values = c(greatersyd_colour,wholensw_colour)) +
  theme(
    axis.text.x=element_text(angle = 45, vjust=1,hjust=1),
    plot.title = element_text(hjust = 0.5)
    ) +
  labs(x = element_blank(),y = element_blank(),colour = element_blank(),
       title="Total Number of Rental Bonds in Greater Sydney")

fig_comp
```

The quarter ending March 2023 does mark a change in the trend of rental bonds in the Greater Sydney region, albeit not as drastic as the Inner West LGA.

The ring classification was introduced by the Bureau of Communications, Arts and Regional Research (BCARR) for dividing Australian cities. The inner ring contains a cities CBD and inner suburbs, of which the Inner West LGA belongs to. The middle ring contains lower density housing but have some high density areas, examples of middle ring LGA's include Canterbury-Bankstown and Parramatta. The outer ring refers to the outer limits of a city, examples include Blacktown and Hornsby LGA.

Pre-covid all three regions were on a steady trend up. During the pandemic, the inner and middle rings took a dip towards the start of the period, but regained traction shortly after. The outer ring maintained it's upward trend through the pandemic. Post-pandemic, the trend for inner and middle ring rental bonds is unsteady, with both losing a few thousand bonds between June 2023 and June 2024 (3,088 lower for the inner ring and 2,580 lower for the middle ring). The outer ring on the other hand added 1,729 bonds in that same period.

### Number of Rental Bonds by Number of Bedrooms and Dwelling Type

```{r}
#| label: fig-iw-bonds-dwelling-type

fig_iw_bonds_bedrooms <- inner_west_rents |>
  dplyr::filter(dwelling_type == "Total" & no_of_bedrooms != "Total" & no_of_bedrooms != "Not Specified" & date == "2024-06-30") |> 
  ggplot() +
  aes(x = factor(no_of_bedrooms),y=no_of_bonds_total,fill=factor(no_of_bedrooms)) +
  scale_y_continuous(labels=scales::comma) +
  scale_x_discrete(labels = c("4 or more Bedrooms" = "4 or more\nBedrooms")) +
  geom_bar(stat="identity") +
  labs(fill=element_blank(),
       x = element_blank(),
       y= element_blank(),
       title="Total Number of Bonds\nby Number of Bedrooms") +
  theme(axis.text.x=element_text(angle = 45, vjust=1,hjust=1),plot.title = element_text(hjust = 0.5),legend.position="none")
```

Rental bonds for houses and flats are shown to have an upward trend over the period, whereas town houses and other dwelling types have lost a share of bonds.

```{r}

fig_iw_bonds_dwelling <- inner_west_rents |>
  dplyr::filter(dwelling_type != "Total" & no_of_bedrooms == "Total" & date == "2024-06-30") |> 
  ggplot() +
  aes(x = dwelling_type,y=no_of_bonds_total,fill=dwelling_type) +
  geom_bar(stat="identity") +
  scale_y_continuous(labels=scales::comma) +
  labs(fill=element_blank(),
       x = element_blank(),
       y= element_blank(),
       title="Total Number of Bonds\nby Dwelling Type") +
  theme(plot.title = element_text(hjust = 0.5),legend.position = "none")

ggarrange(fig_iw_bonds_bedrooms,fig_iw_bonds_dwelling,ncol=2,labels=c("i.","ii."))
#fig_iw_bonds_dwelling
```

## Property Sales in the Inner West

```{r}
#| label: fig-inner-west-sales
#| fig-cap: "Median property sales price of the Inner West LGA"

fig_sales <- sales_df |>
  dplyr::filter(
    ((lga == "Inner West" & greater_sydney == "Total") |
       (lga == "Total" & greater_sydney == "Greater Sydney")) &
      gmr == "Total" & rings == "Total" & dwelling_type == "Total"
  ) |>
  ggplot() +
  aes(x = date, y = median_sales_price,colour = factor(lga)) +
  labs(x = element_blank(),
       y = "Sales Price",
       title = "Median Sales Price of the Inner West\nand Greater Sydney",
       colour = element_blank()) +
  geom_point() +
  geom_line(linetype = "dashed") +
  scale_x_date(breaks = scale ,date_labels = "%b %Y") +
  scale_y_continuous(labels=scales::dollar_format()) +
  geom_vline(xintercept=as.Date("2020-03-15"),linetype="dotted") +
  geom_vline(xintercept=as.Date("2021-10-11"),linetype="dotted") +
  annotate("text",label="Pandemic\nPeriod",x=as.Date("2021-01-01"),y=1550) +
  scale_color_manual(values = c(innerwest_colour,greatersyd_colour,wholensw_colour),
                     labels = c("Inner West","Greater Sydney")) +
  theme(
    axis.text.x=element_text(angle = 45, vjust=1,hjust=1),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    legend.key = element_rect(fill = "white")
    )

fig_sales

```

As shown in @fig-inner-west-sales, the median property price in the Inner West peaked in the quarter ending December 2021, after which it took a downturn over the succeeding three quarters. From the quarter ending September 2022, there was consistent growth for five quarters, until the last reported quarter ending March 2024, where the median property price held steady.

The median property price in the Inner West LGA consistently sits above the median property price of Greater Sydney across all quarters in the reporting period. In the quarter ending March 2024, the last quarter reported, the Inner West LGA had a median property price of \$1,500,000, \$1,000,000 for the Greater Sydney region, and \$870,000 for NSW.

## The Effect of the RBA Cash Rate on Rents and Sales in the Inner West LGA

```{r}
#| label: fig-rba-cr
#| fig-cap: "Reserve Bank of Australia overnight cash rate over the period this report is looking at"

fig_rba_cr <- RBA_cash_rate |>
  dplyr::filter(date > as.Date("2017-12-31") & date < as.Date("2024-07-01")) |>
  ggplot() +
  aes(x = date,y = cash_rate) +
  geom_line() +
  scale_x_date(breaks = scale,date_labels = "%b %Y") +
  scale_y_continuous(labels=scales::label_percent(scale=1)) +
  geom_vline(xintercept=as.Date("2020-03-15"),linetype="dotted") +
  geom_vline(xintercept=as.Date("2021-10-11"),linetype="dotted") +
  annotate("text",label="Pandemic\nPeriod",x=as.Date("2021-01-01"),y=3) +
  theme(axis.text.x=element_text(angle = 45, vjust=1,hjust=1),
    plot.title = element_text(hjust = 0.5)) +
  labs(x = element_blank(), y = "Overnight Cash Rate",
       title = "Reserve Bank of Australia Overnight Cash Rate")
fig_rba_cr
```

The RBA overnight cash rate has changed considerably over the period this report looks at. It hit historic lows during the pandemic, hitting a minimum of 0.10% (a historic minimum). It currently sits at 4.35%, the highest it has been since November 2011. Within the span of two and a half years, the cash rate has risen 4.25% from it's minimum. 

```{r}
#| label: fig-iw-rents-against-cr
#| fig-cap: "Median rent in the Inner West LGA against the RBA cash rate"

fig_iw_rent_cr <- inner_west_rents |> 
  dplyr::filter(dwelling_type == "Total" & no_of_bedrooms == "Total") |>
  ggplot() +
  aes(x = cash_rate, y = median_rent_price) +
  geom_point(colour = innerwest_colour) +
  scale_y_continuous(labels=scales::dollar_format()) +
  scale_x_continuous(breaks=seq(0,4.5,by=0.5)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "Cash Rate",y = "Median Rent",title = "Median Rent of the Inner West LGA\nagainst RBA Cash Rate")
fig_iw_rent_cr
```

The median rent in the Inner West against cash rate is strongly correlated, with a Pearson correlation coefficient of 0.96. The correlation between total number of bonds held and cash rate isn't as strong, with a correlation coefficient of 0.63. Correlation between median sales price and RBA cash rate is relatively weak, with a correlation coefficient of 0.18. Number of properties sold is also weakly correlated with a correlation coefficient of -0.13

## Conclusion
